The Chemistry Development Kit (CDK). 3. Atom typing, Rendering, Molecular Formula, and Substructure Searching

### 2014-07-03 article.tex

\begin{abstract}
% Do not use inserted blank lines (ie \\) until main body of text.
\paragraph*{Background:}
Cheminformatics is a well-established field with many applications in chemistry,
biology, drug discovery, and others. The Chemistry Development Kit (CDK) has
become a widely used Open Source cheminformatics toolkit, providing various models to represent
chemical structures, of which the chemical graph is essential. However,
in the first five years of the project increased so much in size that
interdependencies between components grew unmanageable large, resulting
in unpredictable instabilities.
\paragraph*{Results:}
We here report the improvements of the CDK since the 1.2 release series made
to accomodate both the increased complexity of the library, as well as
siginificant improvements of and additions to the functionality of the
library. We here outline how the CDK evolved since the previous report and
discuss the quality control approached we have adopted to ensure
stability. Additionally, new APIs have been introduced for atom type
perception, substructure searching, rendering of molecules, and handling of
molecular formulas.
\paragraph*{Conclusions:}
With this paper we have shown the continued effort to provide a free, Open Source
cheminformatics library, and show that such colleborative projects can survive for
a long period. We have taken advantage from the community support, and show
that an open source cheminformatics project can act as a peer reviewed
publishing platform for scientific computating software.
\end{abstract}

@@

Open Source cheminformatics has made significants steps forward recently~\cite{OBoyle2011}.
The Chemistry Development Kit (CDK) is one of the tools under the Blue Obelisk wings,
and previously documented version have been widely adopted~\cite{Steinbeck2003,Steinbeck2006}.
For example, new tools expose CDK functionality, such as Cinfony~\cite{OBoyle2008},
rcdk~\cite{}, and workflow plugins for Taverna\cite{Truszkowski2011} and KNIME\cite{Beisken2013}.
Furthermore, projects with specific functionality have emerged building on the CDK, including
jCompoundMapper~\cite{Hinselmann2011}, ScaffoldHunter~\cite{wetzel2009interactive}, OMG~\cite{Peironcely2012},
Padel~\cite{yap2011padel}, AMBIT-SMARTS~\cite{jeliazkova2011ambitsmarts}, AMBIT-Tautomer~\cite{kochev2013ambit},
and SMSD~\cite{}. Other tools used previous versions already, and follow
newer releases of the CDK, such as Bioclipse~\cite{} and AMBIT~\cite{jeliazkova2011ambit}.

@@

\subsubsection*{Molecular fingerprints}
Also the molecular fingerprints in CDK has seen development since the 1.2
release. Traditionally CDK fingerprints were represented using the
\textit{BitSet} class shipping with Java. The benefits of the class include
such things as already implemented methods for modifying bit sets using other
bit sets. However the class keeps a vector of bits in memory. The solution was
excellent for hashed, relatively small fingerprints, \textit{e.g.}, 1024 bits,
\textit{i.e.}, 10 bits indexing space. However implementing a fingerprint
designed to avoid collisions with, \textit{e.g.} a 32 bit indexing space using
this approach would be highly memory-inefficient. To allow for multiple
fingerprint representation implementations a bit fingerprint interface was
introduced. Also, although fingerprints traditionally are bit vectors a count
fingerprint was also introduced making fingerprints based on integer vectors
supported in CDK as well. The fingerprints currently existing in CDK are listed
in Table~\ref{tab:fingerprints}.

@@

%% Use of \listoftables is discouraged.
%%
\section*{Tables}
  \subsection*{Table 1 - The molecular fingerprints in CDK}
    \rowcolors{1}{}{lightgray}
    \par
    \mbox{
    \begin{tabular}{lcclrc}
    \toprule
                             & Bit version  & Count version & CDK version & Default Size & References \\
  EStateFingerprinter        & $\checkmark$ &               & 1.2.0       & $79$         & \cite{Hall1995}\\
  ExtendedFingerprinter      & $\checkmark$ &               & 1.0         &              & \\
  Fingerprinter              & $\checkmark$ &               & 1.0         &              & \\
  GraphOnlyFingerprinter     & $\checkmark$ &               & 1.0         &              & \\
  HybridizationFingerprinter & $\checkmark$ &               & 1.4.0       &              & \\
  KlekotaRothFingerprinter   & $\checkmark$ &               & 1.4.6       & $4\,860$     & \cite{Klekota2008}\\
  LingoFingerprinter         & $\checkmark$ &               & 1.6.0 ?     &              & \\
  MACCSFingerprinter         & $\checkmark$ &               & 1.2.0       &              & \\
  PubchemFingerprinter       & $\checkmark$ &               & 1.4.0       &              & \\
  ShortestPathFingerprinter  & $\checkmark$ &               & 1.6.0 ?     &              & \\
  SignatureFingerprinter     & $\checkmark$ & $\checkmark$  & 1.6.0 ?     & $2^{32}$     & [pending] \\
  SubstructureFingerprinter  & $\checkmark$ &               & 1.0         &              & \\
  CirkularFingerprinter      & $\checkmark$ & ?             & ?           & ?            & [pending?] \\
  \bottomrule
    \end{tabular}}

>>>>

\begin{abstract}
% Do not use inserted blank lines (ie \\) until main body of text.
\paragraph*{Background:}
Cheminformatics is a well-established field with many applications in chemistry,
biology, drug discovery, and others. The Chemistry Development Kit (CDK) has
become a widely used Open Source cheminformatics toolkit, providing various models to represent
chemical structures, of which the chemical graph is essential. However,
in the first five years of the project increased so much in size that
interdependencies between components grew unmanageable large, resulting
in unpredictable instabilities.
\paragraph*{Results:}
We here report the improvements of the CDK since the 1.2 release series made
to accomodate both the increased complexity of the library, as well as
siginificant improvements of and additions to the functionality of the
library. We here outline how the CDK evolved since the previous report and
discuss the quality control approached we have adopted to ensure stability.
Additionally, new APIs have been introduced for atom type perception,
substructure searching, molecular fingerprints, rendering of molecules, and
handling of molecular formulas.
\paragraph*{Conclusions:}
With this paper we have shown the continued effort to provide a free, Open Source
cheminformatics library, and show that such colleborative projects can survive for
a long period. We have taken advantage from the community support, and show
that an open source cheminformatics project can act as a peer reviewed
publishing platform for scientific computing software.
\end{abstract}

@@

Open Source cheminformatics has made significants steps forward recently~\cite{OBoyle2011}.
The Chemistry Development Kit (CDK) is one of the tools under the Blue Obelisk wings,
and previously documented version have been widely adopted~\cite{Steinbeck2003,Steinbeck2006}.
For example, new tools expose CDK functionality, such as Cinfony~\cite{OBoyle2008},
rcdk~\cite{}, and workflow plugins for Taverna\cite{Truszkowski2011} and KNIME\cite{Beisken2013}.
Furthermore, projects with specific functionality have emerged building on the CDK, including
jCompoundMapper~\cite{Hinselmann2011}, ScaffoldHunter~\cite{wetzel2009interactive}, OMG~\cite{Peironcely2012},
Padel~\cite{yap2011padel}, AMBIT-SMARTS~\cite{jeliazkova2011ambitsmarts}, AMBIT-Tautomer~\cite{kochev2013ambit},
and SMSD~\cite{}. Other tools used previous versions already, and follow
newer releases of the CDK, such as Bioclipse~\cite{spjuth2007bioclipse,
spjuth2009bioclipse} and AMBIT~\cite{jeliazkova2011ambit}.

@@

\subsubsection*{Molecular fingerprints}
Also the molecular fingerprints in CDK has seen development since the 1.2
release. Traditionally CDK fingerprints were represented using the
\textit{BitSet} class shipping with Java. The benefits of that class include
such things as already implemented methods for modifying bit sets using other
bit sets. However the class keeps a vector of bits in memory. The solution was
excellent for hashed, relatively small fingerprints, \textit{e.g.}, 1024 bits,
\textit{i.e.}, 10 bits indexing space. However implementing a fingerprint
designed to avoid collisions with, \textit{e.g.} a 32 bit indexing space using
this approach would be highly memory-inefficient. To allow for multiple
fingerprint representation implementations a bit fingerprint interface was
introduced. Also, although fingerprints traditionally are bit vectors a count
fingerprint was also introduced making fingerprints based on integer vectors
supported in CDK as well. The fingerprints currently existing in CDK are listed
in Table~\ref{tab:fingerprints}.

@@

%% Use of \listoftables is discouraged.
%%
\section*{Tables}
  \subsection*{Table 1 - The molecular fingerprints in CDK}
    \rowcolors{1}{}{lightgray}
    \par
    \mbox{
    \begin{tabular}{lcclrc}
    \toprule
                             & Bit version  & Count version & CDK version & Default Size & References \\
  CirkularFingerprinter      & $\checkmark$ & ?             & ?           & ?            & [pending?] \\
  EStateFingerprinter        & $\checkmark$ &               & 1.2.0       & $79$         & \cite{Hall1995}\\
  ExtendedFingerprinter      & $\checkmark$ &               & 1.0         &              & \\
  Fingerprinter              & $\checkmark$ &               & 1.0         &              & \\
  GraphOnlyFingerprinter     & $\checkmark$ &               & 1.0         &              & \\
  HybridizationFingerprinter & $\checkmark$ &               & 1.4.0       &              & \\
  KlekotaRothFingerprinter   & $\checkmark$ &               & 1.4.6       & $4\,860$     & \cite{Klekota2008}\\
  LingoFingerprinter         & $\checkmark$ &               & 1.6.0 ?     &              & \\
  MACCSFingerprinter         & $\checkmark$ &               & 1.2.0       &              & \\
  PubchemFingerprinter       & $\checkmark$ &               & 1.4.0       &              & \\
  ShortestPathFingerprinter  & $\checkmark$ &               & 1.6.0 ?     &              & \\
  SignatureFingerprinter     & $\checkmark$ & $\checkmark$  & 1.6.0 ?     & $2^{32}$     & [pending] \\
  SubstructureFingerprinter  & $\checkmark$ &               & 1.0         &              & \\
  \bottomrule
    \end{tabular}}
